Dark Control:\\ Towards a Unified Account of Default Mode Function\\ by Markov Decision Processes

### 2017-12-07 paper/paper.tex
a

\begin{abstract}
The Default Mode Network (DMN) is believed to subserve the
baseline mental activity in humans.
%
Its highest energy consumption compared to other brain networks and
its intimate coupling with conscious awareness are both pointing to
an overarching function.
%
Many research streams speak in favor of an evolutionarily adaptive role in
envisioning experience to anticipate the future.
In the present work, we propose a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
% DMN function is recast in terms of
% control theory and reinforcement learning by basing it on Markov decision processes.
%
Specifically, we answer the question whether the neurobiological properties of the default mode network collectively provide the computational building blocks necessary for a Markov Decision Process.
%
We argue that our formal account of DMN function naturally accommodates as special cases
the previously proposed cognitive accounts on
(1) predictive coding,
(2) semantic associations, and
(3) a ``sentinel'' role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

Such an RL account of DMN function can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term gratification.

We argue that DMN implication in many
advanced cognitive processes in humans
can be recast as prediction error minimization
based on probabilistic mental simulations
- ``covert forms of action and perception'' \citep{pezzulo2011grounding} -,
thus maximizing action outcomes across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling.
Even (unavoidably) imperfect memory recall,
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide randomly sampled blocks of pseudo-experience instrumental to iteratively
optimize the behavior of the organism.

@@

\subsection{Putting everything together}
The DMN is today known to consistently increase in neural
activity when humans engage in cognitive processes that are relatively detached from
the current sensory environment.

The more familiar and predictable the current environment, the more resources remain for allocating DMN activity to MDP processes extending beyond the present time and sensory context. In line with this perspective, DMN engagement was shown to heighten and relate to effective behavioral responses in the practiced phase of a demanding cognitive flexibility task, as compared to acquisition phase when participants learned context-specific rules. This involvement in automated decision-making lead the authors to propose an ``autopilot'' role for the DMN \citep{vatansever2017pnas}, which may contribute to optimizing control of the organism in general. Among all DMN nodes, the RTPJ has probably a most evident role as network-switching relay calibrating between processing of environment-engaged versus internally generated information \citep{downar2000multimodal, golland2006extrinsic, bzdok2013tpj}.

Additionally,
this network was proposed to be situated
at the top of the brain network hierarchy, with
the subordinate salience and dorsal attention network in the middle and
the primary sensory cortices at the bottom
\citep{carhart2010default, margulies2016situating}.
Its putative involvement in thinking about the past,
hypothetical experiences, and the future
appears to tie in with the implicit computation of
action and state cascades as a function of
experienced events and collected feedback from the past.

A policy matrix encapsulates the repertoire of possible actions
on the world given a current state.
The policy matrix encodes the probabilites of
choosing actions to be executed in a certain situation.
The DMN may subserve
  constant exploration of possible action trajectories and
  nested estimaton of their
  cumulative reward outcomes. Implicit computation of future choices
  provides a potential explanation for the
  evolutionary emergence and practical usefulness of
  mind-wandering at day-time and dreams during sleep
  in humans.
\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{neurobiological_and_rl_overview_DMN.pdf}
  \caption{\textbf{Default mode network:
  neurobiological implementation of reinforcement learning.}
  Overview of how the constituent nodes of the DMN (refer to section \ref{sec:nodes}) may
  map onto computational components necessary for an RL agent.}
  \label{fig:rl_process_chart}
\end{figure}

The HC may contribute to generation of
perturbed action-transition-state-reward samples
as batches of pseudo-experience
(i.e., imagined, hypothesized, and recalled mental scenarios).
The small variations in these experience samplings allow searching
a larger space of model parameters and possible experiences.
Taken to its extreme, stochastic recombination of experience
building blocks can further optimize the behavior of the RL agent
by model learning from scenarios in the environment that the agent might
only very rarely or never encounter.
An explanation is thus offered for experiencing seemingly familiar situations that
a human has however never actually encountered (i.e., d\'{e}j\`{a} vu effect).
While such a situation may not have been experienced in the physical world,
the DMN may have previously stochastically generated, evaluated, and adapted to
such a randomly synthesized situation.
Generated representations arguably are ``internally manipulable,
and can be used for attempting actions internally,
before or instead of acting in the external reality,
and in diverse goal and sensory contexts, i.e. even outside the context in which they were learned''
\citep{pezzulo2011grounding}.
In the absence of environmental input and feedback
(e.g., mind-wandering or sleep),
mental scene construction allows for pseudo-experiencing possible
future scenarios and action outcomes.
Our formal account of DMN function
thus acknowledges the unavoidable stochasticity of
computation in neural systems \citep{faisal2008noise}.

From the perspective of a model-free RL agent,
\textit{prediction} in the human brain reduces to
generalization of
policy and value computations from sampled experiences to
successful action choices and reward predictions in future states.
As such,
plasticity in the DMN arises naturally.
If an agent behaving optimally in a certain environment moves
to new, yet unexperienced environment, reward prediction errors will
largely increase.
This feedback will lead to adaptation of policy considerations
and value estimations until the intelligent system converges to a
new steady state of optimal action decisions in a volatile world.

@@

\begin{mdframed}
  \vspace{1em}
  \textbf{Box 2: Hypotheses for testing the MDP account of DMN function}
\begin{enumerate}
\item Experiment (Humans): We hypothesize a functional relationship between the DMN closely associated with the occurrence of stimulus-independent thoughts and the reward circuitry. During an iterative neuroeconomic two-player game, fMRI signals in the DMN could be used to predict reward-related signals in the nucleus accumbens across trials in a continuous learning paradigm. We expect that the more DMN activity is measured to be increased, supposedly the higher the tendency for stimulus-independent thoughts, and the more the fMRI signals in the reward circuits should be independent of the reward context
in the sensory environment.

>>>>

\begin{abstract}
The Default Mode Network (DMN) is believed to subserve the
baseline mental activity in humans.
%
Its highest energy consumption compared to other brain networks and
its intimate coupling with conscious awareness are both pointing to
an overarching function.
%
Many research streams speak in favor of an evolutionarily adaptive role in
envisioning experience to anticipate the future.
In the present work, we propose a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
% DMN function is recast in terms of
% control theory and reinforcement learning by basing it on Markov decision processes.
%
Specifically, we answer the question whether the neurobiological properties of the DMN
collectively provide the computational building blocks necessary for a Markov Decision Process.
%
We argue that our formal account of DMN function naturally accommodates as special cases
the previously proposed cognitive accounts on
(1) predictive coding,
(2) semantic associations, and
(3) a ``sentinel'' role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

Such an RL account of DMN function can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term gratification.
We argue that DMN implication in many
advanced cognitive processes in humans
can be recast as prediction error minimization
based on probabilistic mental simulations
- ``covert forms of action and perception'' \citep{pezzulo2011grounding} -,
thus maximizing action outcomes across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling.
Even (unavoidably) imperfect memory recall,
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide randomly sampled blocks of pseudo-experience instrumental to iteratively
optimize the behavior of the organism.

@@

\subsection{Putting everything together}
The DMN is today known to consistently increase in neural
activity when humans engage in cognitive processes that are relatively detached from
the current sensory environment.
The more familiar and predictable the current environment, the more brain resources may remain for allocating DMN activity to MDP processes extending beyond the present time and sensory context. In line with this perspective, DMN engagement was shown to heighten and relate to effective behavioral responses in the practiced phase of a demanding cognitive flexibility task, as compared to acquisition phase when participants learned context-specific rules. This involvement in automated decision-making has led the authors to propose an ``autopilot'' role for the DMN \citep{vatansever2017pnas}, which may contribute to optimizing control of the organism in general. Among all DMN nodes, the RTPJ is perhaps the most evident candidate for a network-switching relay that calibrates between processing of environment-engaged versus internally generated information \citep{downar2000multimodal, golland2006extrinsic, bzdok2013tpj}.

Additionally,
the DMN was proposed to be situated
at the top of the brain network hierarchy, with
the subordinate salience and dorsal attention network in the middle and
the primary sensory cortices at the bottom
\citep{carhart2010default, margulies2016situating}.
Its putative involvement in thinking about 
hypothetical experiences and future outcomes
appears to tie in with the implicit computation of
action and state cascades as a function of
experienced events and collected feedback from the past.
A policy matrix encapsulates the choice probabilities of action repertoire
on the world given a current situation (i.e., state).
The DMN may subserve
  constant exploration of possible action trajectories and
  nested estimaton of their
  cumulative reward outcomes. Implicit computation of future choices
  provides a potential explanation for the
  evolutionary emergence and practical usefulness of
  mind-wandering at day-time and dreams during sleep
  in humans.
\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{neurobiological_and_rl_overview_DMN.pdf}
  \caption{\textbf{Default mode network:
  neurobiological implementation of reinforcement learning.}
  Overview of how the constituent nodes of the DMN (refer to section \ref{sec:nodes}) may
  map onto computational components necessary for an RL agent.}
  \label{fig:rl_process_chart}
\end{figure}

The HC may contribute to generation of
perturbed action-transition-state-reward samples
as batches of pseudo-experience
(i.e., recalled, hypothesized, and forecasted mental scenarios).
The small variations in these experience samplings allow searching
a larger space of model parameters and possible experiences.
Taken to its extreme, stochastic recombination of experience
building blocks can further optimize the behavior of the RL agent
by model learning from scenarios in the environment that the agent might
only very rarely or never encounter.
An explanation is thus offered for experiencing seemingly familiar situations that
a human has however never actually encountered (i.e., d\'{e}j\`{a} vu effect).
While such a situation may not have been experienced in the physical world,
the DMN may have previously stochastically generated, evaluated, and adapted to
such a randomly synthesized event.
Generated representations arguably are ``internally manipulable,
and can be used for attempting actions internally,
before or instead of acting in the external reality,
and in diverse goal and sensory contexts, i.e. even outside the context in which they were learned''
\citep{pezzulo2011grounding}.
In the context of scarce environmental input and feedback
(e.g., mind-wandering or sleep),
mental scene construction allows pseudo-experiencing possible
future scenarios and action outcomes.

From the perspective of a model-free RL agent,
prediction in the DMN reduces to
generalization of
policy and value computations from sampled experiences to
successful action choices and reward predictions in future states.
As such,
plasticity in the DMN arises naturally.
If an agent behaving optimally in a certain environment moves
to new, yet unexperienced environment, reward prediction errors will
largely increase.
This feedback will lead to adaptation of policy considerations
and value estimations until the intelligent system converges to a
new steady state of optimal action decisions in a volatile world.

@@

\begin{mdframed}
  \vspace{1em}
  \textbf{Box 2: Future studies for testing the MDP account of DMN function}
\begin{enumerate}
\item Experiment (Humans): We hypothesize a functional relationship between the DMN closely associated with the occurrence of stimulus-independent thoughts and the reward circuitry. During an iterative neuroeconomic two-player game, fMRI signals in the DMN could be used to predict reward-related signals in the nucleus accumbens across trials in a continuous learning paradigm. We expect that the more DMN activity is measured to be increased, supposedly the higher the tendency for stimulus-independent thoughts, and the more the fMRI signals in the reward circuits should be independent of the reward context
in the sensory environment.
